---
title: "01 — Exploratory Data Analysis"
format: html
execute:
  message: false
  warning: false
---

```{r setup, include=FALSE}
root_dir <- if (file.exists("data/raw")) "." else ".."
knitr::opts_knit$set(root.dir = normalizePath(root_dir))
```


## Load raw hospital dataset

```{r load-data}
library(tidyverse)

fig_dir <- "output/figures"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
fill_colors <- c("No" = "#4C78A8", "Yes" = "#E45756")

csv_path <- "data/raw/diabetic_data.csv"
if (!file.exists(csv_path)) {
  stop("Dataset not found. Run source('scripts/download_data.R') first.")
}

raw <- read_csv(csv_path, show_col_types = FALSE)
raw <- raw |>
  mutate(across(where(is.character), ~ na_if(.x, "?")))

mapping_path <- "data/raw/IDS_mapping.csv"
if (!file.exists(mapping_path)) {
  stop("ID mapping file not found. Re-run source('scripts/download_data.R').")
}

load_id_mapping <- function(section) {
  lines <- readLines(mapping_path)
  header_idx <- which(lines == paste0(section, ",description"))[1]
  comma_idxs <- which(lines == ",")
  stop_idx <- comma_idxs[comma_idxs > header_idx][1]
  body <- lines[(header_idx + 1):(stop_idx - 1)]
  df <- read_csv(
    I(c(paste0(section, ",description"), body)),
    show_col_types = FALSE
  )
  setNames(df$description, as.character(df[[section]]))
}

admission_type_labels <- load_id_mapping("admission_type_id")
discharge_disposition_labels <- load_id_mapping("discharge_disposition_id")
```


## Define 30-day readmission target

```{r target-variable}
age_levels <- c(
  "[0-10)", "[10-20)", "[20-30)", "[30-40)", "[40-50)",
  "[50-60)", "[60-70)", "[70-80)", "[80-90)", "[90-100)"
)

eda_data <- raw |>
  mutate(
    readmit_30 = case_when(
      readmitted == "<30" ~ "Yes",
      readmitted %in% c("NO", ">30") ~ "No",
      TRUE ~ NA_character_
    ),
    readmit_30 = factor(readmit_30, levels = c("No", "Yes")),
    age = factor(age, levels = age_levels)
  ) |>
  filter(!is.na(readmit_30))

cat("Observations:", nrow(eda_data), "\n")
cat("30-day readmission rate:",
    scales::percent(mean(eda_data$readmit_30 == "Yes"), accuracy = 0.1), "\n")
```


## Explore class imbalance in readmission

```{r class-imbalance}
class_counts <- eda_data |>
  count(readmit_30) |>
  mutate(pct = n / sum(n))

class_counts

p_imbalance <- ggplot(class_counts, aes(x = readmit_30, y = pct, fill = readmit_30)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Class Distribution (30-Day Readmission)", x = NULL, y = "Proportion") +
  theme_minimal()

print(p_imbalance)
```


## Readmission by age and utilization

```{r age-utilization}
#| fig-width: 8
#| fig-height: 5
#| fig-show: hold

p_age <- ggplot(eda_data, aes(x = age, fill = readmit_30)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Readmission Rate by Age Group", x = "Age group", y = "Proportion") +
  theme_minimal() +
  theme(legend.title = element_blank())

p_los <- ggplot(eda_data, aes(x = time_in_hospital, fill = readmit_30)) +
  geom_histogram(binwidth = 1, position = "fill", color = "white") +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Readmission Rate by Length of Stay", x = "Days in hospital", y = "Proportion") +
  theme_minimal() +
  theme(legend.title = element_blank())

p_meds <- ggplot(eda_data, aes(x = num_medications, fill = readmit_30)) +
  geom_histogram(binwidth = 1, position = "fill", color = "white") +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Readmission Rate by Number of Medications", x = "Number of medications", y = "Proportion") +
  theme_minimal() +
  theme(legend.title = element_blank())

print(p_age)
print(p_los)
print(p_meds)
```


## Summary statistics for numeric predictors

```{r summary-stats}
eda_data |>
  select(time_in_hospital, num_lab_procedures, num_medications, number_diagnoses) |>
  summary()
```


## Missing values by variable

```{r missing-values}
missing_pct <- eda_data |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing))

missing_pct
```


## Identify low-information variables to exclude

```{r variables-to-exclude}
low_info <- tibble(
  variable = c("weight", "max_glu_serum", "A1Cresult"),
  pct_missing_or_unmeasured = c(
    mean(is.na(eda_data$weight)),
    mean(eda_data$max_glu_serum == "None", na.rm = TRUE),
    mean(eda_data$A1Cresult == "None", na.rm = TRUE)
  ),
  recommendation = "Exclude from modeling"
) |>
  mutate(pct_missing_or_unmeasured = scales::percent(pct_missing_or_unmeasured, accuracy = 0.1))

low_info
```


## Categorical predictor distributions

```{r categorical}
cat("Raw readmission status:\n")
print(table(raw$readmitted, useNA = "ifany"))

cat("\nAge group counts:\n")
print(table(eda_data$age))
```


## Readmission by clinical factors

```{r clinical-factors}
#| fig-width: 8
#| fig-height: 6
#| fig-show: hold

label_id <- function(x, labels) {
  mapped <- labels[as.character(x)]
  ifelse(is.na(mapped), paste("ID", x), mapped)
}

plot_readmit_rate <- function(data, x_var, title, horizontal = FALSE) {
  data <- data |>
    mutate(
      !!sym(x_var) := fct_reorder(
        .data[[x_var]],
        readmit_30,
        .fun = function(x) mean(x == "Yes")
      )
    )

  if (horizontal) {
    ggplot(data, aes(y = .data[[x_var]], fill = readmit_30)) +
      geom_bar(position = "fill") +
      scale_fill_manual(values = fill_colors) +
      scale_y_discrete(labels = function(x) stringr::str_wrap(x, 40)) +
      labs(title = title, x = "Proportion", y = NULL) +
      coord_cartesian(clip = "off") +
      theme_minimal() +
      theme(
        legend.title = element_blank(),
        axis.text.y = element_text(size = 8),
        plot.margin = margin(10, 10, 10, 10)
      )
  } else {
    ggplot(data, aes(x = .data[[x_var]], fill = readmit_30)) +
      geom_bar(position = "fill") +
      scale_fill_manual(values = fill_colors) +
      labs(title = title, x = NULL, y = "Proportion") +
      coord_cartesian(clip = "off") +
      theme_minimal() +
      theme(
        legend.title = element_blank(),
        plot.margin = margin(10, 10, 10, 10)
      )
  }
}

eda_labeled <- eda_data |>
  mutate(
    admission_type = label_id(admission_type_id, admission_type_labels),
    discharge_disposition = label_id(discharge_disposition_id, discharge_disposition_labels)
  )

p_admission <- plot_readmit_rate(
  eda_labeled,
  "admission_type",
  "Readmission Rate by Admission Type",
  horizontal = TRUE
)

p_discharge <- plot_readmit_rate(
  eda_labeled,
  "discharge_disposition",
  "Readmission Rate by Discharge Disposition",
  horizontal = TRUE
)

p_diabetes_med <- plot_readmit_rate(eda_data, "diabetesMed", "Readmission Rate by Diabetes Medication")

p_change <- plot_readmit_rate(eda_data, "change", "Readmission Rate by Medication Change")

print(p_admission)
print(p_diabetes_med)
print(p_change)
```

```{r discharge-disposition}
#| fig-width: 8
#| fig-height: 12

print(p_discharge)
```


## Death and hospice discharge dispositions

```{r death-hospice}
death_hospice_ids <- c(11, 13, 14, 19, 20, 21)

death_hospice <- eda_data |>
  filter(discharge_disposition_id %in% death_hospice_ids) |>
  mutate(
    disposition = discharge_disposition_labels[as.character(discharge_disposition_id)]
  ) |>
  count(disposition, readmit_30) |>
  pivot_wider(names_from = readmit_30, values_from = n, values_fill = 0) |>
  mutate(encounters = No + Yes, pct_of_data = scales::percent(encounters / nrow(eda_data), accuracy = 0.1)) |>
  select(disposition, encounters, pct_of_data, No, Yes)

death_hospice
```


## Numeric predictor distributions by readmission

```{r boxplots}
p_box_los <- ggplot(eda_data, aes(x = readmit_30, y = time_in_hospital, fill = readmit_30)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Length of Stay by Readmission Status", x = NULL, y = "Days in hospital") +
  theme_minimal()

p_box_meds <- ggplot(eda_data, aes(x = readmit_30, y = num_medications, fill = readmit_30)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Number of Medications by Readmission Status", x = NULL, y = "Medications") +
  theme_minimal()

print(p_box_los)
print(p_box_meds)
```


## Correlation among predictors

```{r correlation}
numeric_preds <- eda_data |>
  select(
    time_in_hospital, num_lab_procedures, num_procedures, num_medications,
    number_outpatient, number_emergency, number_inpatient, number_diagnoses
  )

var_labels <- c(
  time_in_hospital = "Length of stay",
  num_lab_procedures = "Lab procedures",
  num_procedures = "Procedures",
  num_medications = "Medications",
  number_outpatient = "Outpatient visits",
  number_emergency = "Emergency visits",
  number_inpatient = "Inpatient visits",
  number_diagnoses = "Diagnoses"
)

label_order <- unname(var_labels)

cor_mat <- cor(numeric_preds, use = "pairwise.complete.obs")

cor_long <- cor_mat |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") |>
  mutate(
    var1 = var_labels[var1],
    var2 = var_labels[var2],
    var1 = factor(var1, levels = label_order),
    var2 = factor(var2, levels = rev(label_order))
  )

p_cor <- ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#4C78A8", mid = "white", high = "#E45756",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(title = "Correlation Among Numeric Predictors", x = NULL, y = NULL, fill = "Correlation") +
  coord_fixed() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_cor)
```


## Save EDA figures to output

```{r save-plots}
ggsave(file.path(fig_dir, "01_class_imbalance.png"), p_imbalance, width = 7, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "02_readmission_by_age.png"), p_age, width = 8, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "03_readmission_by_los.png"), p_los, width = 8, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "04_readmission_by_medications.png"), p_meds, width = 8, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "05_readmission_by_admission_type.png"), p_admission, width = 8, height = 6, dpi = 150)
ggsave(file.path(fig_dir, "06_readmission_by_discharge.png"), p_discharge, width = 8, height = 12, dpi = 150)
ggsave(file.path(fig_dir, "07_readmission_by_diabetes_med.png"), p_diabetes_med, width = 7, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "08_readmission_by_change.png"), p_change, width = 7, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "09_boxplot_los.png"), p_box_los, width = 7, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "10_boxplot_medications.png"), p_box_meds, width = 7, height = 5, dpi = 150)
ggsave(file.path(fig_dir, "11_correlation_heatmap.png"), p_cor, width = 9, height = 7, dpi = 150)
```


## Save cleaned data for preprocessing

```{r save-data}
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)
saveRDS(eda_data, "data/processed/eda_data.rds")
```