---
title: "02 — Data Preprocessing"
format: html
execute:
  message: false
  warning: false
---

```{r setup, include=FALSE}
root_dir <- if (file.exists("data/raw")) "." else ".."
knitr::opts_knit$set(root.dir = normalizePath(root_dir))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```



## Load cleaned EDA data

```{r load-eda}
library(tidyverse)
library(caret)

eda_path <- "data/processed/eda_data.rds"
if (!file.exists(eda_path)) {
  stop("EDA data not found. Run R/01_eda.qmd first.")
}

eda_data <- readRDS(eda_path)
```



## Drop IDs, low-information predictors, and death/hospice rows

```{r feature-selection}
death_hospice_ids <- c(11, 13, 14, 19, 20, 21)

drop_cols <- c(
  "encounter_id",
  "patient_nbr",
  "readmitted",
  "weight",
  "max_glu_serum",
  "A1Cresult"
)

model_data <- eda_data |>
  filter(!discharge_disposition_id %in% death_hospice_ids) |>
  select(-any_of(drop_cols)) |>
  mutate(
    readmit_30 = factor(readmit_30, levels = c("No", "Yes")),
    across(where(is.character), as.factor)
  ) |>
  distinct()

cat("Excluded death/hospice rows:", sum(eda_data$discharge_disposition_id %in% death_hospice_ids), "\n")
cat("Modeling rows:", nrow(model_data), "\n")
cat("Predictors:", ncol(model_data) - 1, "\n")
```



## Stratified train validation test split

```{r split}
set.seed(503)

train_val_idx <- createDataPartition(model_data$readmit_30, p = 0.75, list = FALSE)
train_val <- model_data[train_val_idx, ]
test_data <- model_data[-train_val_idx, ]

train_idx <- createDataPartition(train_val$readmit_30, p = 0.8, list = FALSE)
train_data <- train_val[train_idx, ]
val_data <- train_val[-train_idx, ]

split_summary <- tibble(
  split = c("Train", "Validation", "Test"),
  n = c(nrow(train_data), nrow(val_data), nrow(test_data)),
  readmit_rate = c(
    mean(train_data$readmit_30 == "Yes"),
    mean(val_data$readmit_30 == "Yes"),
    mean(test_data$readmit_30 == "Yes")
  )
) |>
  mutate(readmit_rate = scales::percent(readmit_rate, accuracy = 0.1))

split_summary
```



## Impute missing numeric and categorical values

```{r step-impute}
impute_from_train <- function(train, new_data) {
  imputed <- new_data

  for (col in names(train)) {
    if (col == "readmit_30") {
      next
    }

    if (is.numeric(train[[col]])) {
      fill_value <- median(train[[col]], na.rm = TRUE)
    } else {
      fill_value <- names(sort(table(train[[col]]), decreasing = TRUE))[1]
    }

    imputed[[col]][is.na(imputed[[col]])] <- fill_value
  }

  imputed
}

train_imp <- impute_from_train(train_data, train_data)
val_imp <- impute_from_train(train_data, val_data)
test_imp <- impute_from_train(train_data, test_data)
```



## Remove zero-variance predictors

```{r step-zv-1}
remove_zero_var <- function(data) {
  keep <- vapply(
    data,
    function(x) {
      if (is.numeric(x)) {
        var(x, na.rm = TRUE) > 0
      } else {
        length(unique(x[!is.na(x)])) > 1
      }
    },
    logical(1)
  )
  data[, keep, drop = FALSE]
}

train_pred <- train_imp |> select(-readmit_30) |> remove_zero_var()
keep_predictors <- colnames(train_pred)

val_pred <- val_imp |> select(all_of(keep_predictors))
test_pred <- test_imp |> select(all_of(keep_predictors))
```



## Remove highly correlated predictors

```{r step-corr}
numeric_cols <- names(train_pred)[vapply(train_pred, is.numeric, logical(1))]

if (length(numeric_cols) > 1) {
  high_corr <- findCorrelation(
    cor(train_pred[, numeric_cols, drop = FALSE], use = "pairwise.complete.obs"),
    cutoff = 0.9
  )

  if (length(high_corr) > 0) {
    remove_cols <- numeric_cols[high_corr]
    keep_predictors <- setdiff(keep_predictors, remove_cols)
    train_pred <- train_pred[, keep_predictors, drop = FALSE]
    val_pred <- val_pred[, keep_predictors, drop = FALSE]
    test_pred <- test_pred[, keep_predictors, drop = FALSE]
  }
}
```



## One-hot encode categorical predictors

```{r step-dummy}
dummy_vars <- dummyVars(~ ., data = train_pred, fullRank = TRUE)

train_x <- as.data.frame(predict(dummy_vars, train_pred))
val_x <- as.data.frame(predict(dummy_vars, val_pred))
test_x <- as.data.frame(predict(dummy_vars, test_pred))
```



## Remove zero-variance dummy columns

```{r step-zv-2}
non_zero_var <- vapply(train_x, function(x) var(x, na.rm = TRUE) > 0, logical(1))
post_dummy_cols <- colnames(train_x)[non_zero_var]

train_x <- train_x[, post_dummy_cols, drop = FALSE]
val_x <- val_x[, post_dummy_cols, drop = FALSE]
test_x <- test_x[, post_dummy_cols, drop = FALSE]
```



## Center and scale predictors

```{r step-normalize}
pre_process <- preProcess(train_x, method = c("center", "scale"))
```



## Fit preprocessor on training data

```{r prep-preprocessor}
train_processed <- as.data.frame(predict(pre_process, train_x))
```



## Apply preprocessor to train and test

```{r bake-preprocessor}
val_processed <- as.data.frame(predict(pre_process, val_x))
test_processed <- as.data.frame(predict(pre_process, test_x))

baked_train <- bind_cols(readmit_30 = train_imp$readmit_30, train_processed)
baked_val <- bind_cols(readmit_30 = val_imp$readmit_30, val_processed)
baked_test <- bind_cols(readmit_30 = test_imp$readmit_30, test_processed)

cat("Features after preprocessing:", ncol(baked_train) - 1, "\n")
```



## Preprocessing summary

```{r preprocess-summary}
preprocess_summary <- tibble(
  dataset = c("Train (processed)", "Validation (processed)", "Test (processed)"),
  rows = c(nrow(baked_train), nrow(baked_val), nrow(baked_test)),
  features = c(ncol(baked_train) - 1, ncol(baked_val) - 1, ncol(baked_test) - 1)
)

preprocess_summary

cat(
  "\nTest readmission rate (unchanged):",
  scales::percent(mean(test_data$readmit_30 == "Yes"), accuracy = 0.1),
  "\n"
)
```



## Save splits and preprocessor

```{r save}
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)

saveRDS(
  list(
    train = train_data,
    validation = val_data,
    test = test_data,
    impute_from_train = impute_from_train,
    keep_predictors = keep_predictors,
    dummy_vars = dummy_vars,
    post_dummy_cols = post_dummy_cols,
    pre_process = pre_process
  ),
  "data/processed/splits_caret.rds"
)
```
