---
title: "03 — Model Training"
format: html
execute:
  message: false
  warning: false
---

```{r setup, include=FALSE}
root_dir <- if (file.exists("data/raw")) "." else ".."
knitr::opts_knit$set(root.dir = normalizePath(root_dir))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```




## Load preprocessed splits

```{r load-data}
library(tidyverse)
library(caret)
library(yardstick)

splits_path <- "data/processed/splits_caret.rds"
if (!file.exists(splits_path)) {
  stop("Preprocessing output not found. Run R/02_preprocess.qmd first.")
}

prep_obj <- readRDS(splits_path)
train_ref <- prep_obj$train
```




## Prepare feature matrices for train and validation

```{r prepare-features}
prepare_features <- function(train_reference, new_data, prep) {
  imputed <- prep$impute_from_train(train_reference, new_data)
  predictors <- imputed |> select(all_of(prep$keep_predictors))

  features <- as.data.frame(predict(prep$dummy_vars, predictors))
  features <- features[, prep$post_dummy_cols, drop = FALSE]
  processed <- as.data.frame(predict(prep$pre_process, features))

  list(x = processed, y = imputed$readmit_30)
}

val_metrics <- function(fit, x_val, y_val) {
  val_probs <- predict(fit, x_val, type = "prob")[, "Yes"]
  pred_class <- predict(fit, x_val)
  y_factor <- factor(y_val, levels = c("No", "Yes"))

  tibble(
    roc_auc = yardstick::roc_auc_vec(y_factor, val_probs, event_level = "second"),
    sens = yardstick::sensitivity_vec(y_factor, pred_class, event_level = "second"),
    spec = yardstick::specificity_vec(y_factor, pred_class, event_level = "second")
  ) |>
    mutate(across(where(is.numeric), as.numeric))
}

ctrl_none <- trainControl(method = "none", classProbs = TRUE, savePredictions = "final")

tune_on_validation <- function(train_data, x_val, y_val, method, grid, ...) {
  best_auc <- -Inf
  best_fit <- NULL
  results <- vector("list", nrow(grid))

  for (i in seq_len(nrow(grid))) {
    fit_i <- train(
      readmit_30 ~ .,
      data = train_data,
      method = method,
      trControl = ctrl_none,
      tuneGrid = grid[i, , drop = FALSE],
      ...
    )

    metrics_i <- val_metrics(fit_i, x_val, y_val)
    results[[i]] <- bind_cols(grid[i, , drop = FALSE], metrics_i)

    if (metrics_i$roc_auc > best_auc) {
      best_auc <- metrics_i$roc_auc
      best_fit <- fit_i
    }
  }

  list(fit = best_fit, results = bind_rows(results))
}

train_prep <- prepare_features(train_ref, prep_obj$train, prep_obj)
val_prep <- prepare_features(train_ref, prep_obj$validation, prep_obj)

cat("Training features:", ncol(train_prep$x), "\n")
cat("Validation features:", ncol(val_prep$x), "\n")
```




## Downsample training data for class imbalance

```{r downsample}
set.seed(503)

train_balanced <- downSample(
  x = train_prep$x,
  y = train_prep$y,
  yname = "readmit_30"
)

train_balanced |>
  count(readmit_30) |>
  mutate(pct = scales::percent(n / sum(n), accuracy = 0.1))
```




## Logistic regression

```{r logistic}
fit_logistic <- train(
  readmit_30 ~ .,
  data = train_balanced,
  method = "glm",
  family = binomial(),
  trControl = ctrl_none
)

logistic_val <- val_metrics(fit_logistic, val_prep$x, val_prep$y) |>
  mutate(model = "Logistic regression")

logistic_val
```




## glmnet with validation tuning (ridge, elastic net, LASSO)

```{r lasso}
lasso_grid <- expand.grid(
  alpha = c(0, 0.5, 1),
  lambda = seq(0.001, 0.05, length.out = 5)
)

lasso_tuned <- tune_on_validation(
  train_data = train_balanced,
  x_val = val_prep$x,
  y_val = val_prep$y,
  method = "glmnet",
  grid = lasso_grid,
  family = "binomial"
)

fit_lasso <- lasso_tuned$fit
lasso_val <- val_metrics(fit_lasso, val_prep$x, val_prep$y) |>
  mutate(model = "LASSO")

lasso_val
```




## Random forest with validation tuning

```{r random-forest}
set.seed(503)

rf_grid <- expand.grid(
  mtry = c(100, 200),
  splitrule = "gini",
  min.node.size = 5
)

rf_tuned <- tune_on_validation(
  train_data = train_balanced,
  x_val = val_prep$x,
  y_val = val_prep$y,
  method = "ranger",
  grid = rf_grid,
  importance = "none",
  num.trees = 75,
  num.threads = parallel::detectCores()
)

fit_rf <- rf_tuned$fit
rf_val <- val_metrics(fit_rf, val_prep$x, val_prep$y) |>
  mutate(model = "Random forest")

rf_val
```




## XGBoost with validation tuning

```{r xgboost}
library(xgboost)

set.seed(503)

x_train <- train_balanced |> select(-readmit_30) |> as.matrix()
y_train <- ifelse(train_balanced$readmit_30 == "Yes", 1, 0)
x_val_mat <- as.matrix(val_prep$x)
dtrain <- xgb.DMatrix(data = x_train, label = y_train)

xgb_grid <- expand.grid(
  nrounds = c(50, 100),
  max_depth = c(3, 5),
  eta = 0.1,
  gamma = 0,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  subsample = 0.8
)

xgb_metrics <- function(booster, x_val, y_val) {
  probs <- predict(booster, xgb.DMatrix(as.matrix(x_val)))
  y_factor <- factor(y_val, levels = c("No", "Yes"))
  pred_class <- factor(ifelse(probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))

  tibble(
    roc_auc = yardstick::roc_auc_vec(y_factor, probs, event_level = "second"),
    sens = yardstick::sensitivity_vec(y_factor, pred_class, event_level = "second"),
    spec = yardstick::specificity_vec(y_factor, pred_class, event_level = "second")
  ) |>
    mutate(across(where(is.numeric), as.numeric))
}

xgb_results <- vector("list", nrow(xgb_grid))
best_xgb_auc <- -Inf
fit_xgb <- NULL

for (i in seq_len(nrow(xgb_grid))) {
  params_i <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = xgb_grid$max_depth[i],
    eta = xgb_grid$eta[i],
    gamma = xgb_grid$gamma[i],
    colsample_bytree = xgb_grid$colsample_bytree[i],
    min_child_weight = xgb_grid$min_child_weight[i],
    subsample = xgb_grid$subsample[i]
  )

  fit_i <- xgb.train(
    params = params_i,
    data = dtrain,
    nrounds = xgb_grid$nrounds[i],
    verbose = 0
  )

  metrics_i <- xgb_metrics(fit_i, x_val_mat, val_prep$y)
  xgb_results[[i]] <- bind_cols(xgb_grid[i, , drop = FALSE], metrics_i)

  if (metrics_i$roc_auc > best_xgb_auc) {
    best_xgb_auc <- metrics_i$roc_auc
    fit_xgb <- fit_i
  }
}

xgb_tuned <- list(fit = fit_xgb, results = bind_rows(xgb_results))

xgb_val <- xgb_metrics(fit_xgb, x_val_mat, val_prep$y) |>
  mutate(model = "XGBoost")

xgb_val
```




## Compare validation performance

```{r model-comparison}
val_metrics_all <- bind_rows(logistic_val, lasso_val, rf_val, xgb_val) |>
  select(model, roc_auc, sens, spec) |>
  arrange(desc(roc_auc))

val_metrics_all

best_model_name <- val_metrics_all$model[1]
cat("\nBest model on validation:", best_model_name, "\n")
```




## Hyperparameter tuning results

```{r tuning-results}
lasso_tuned$results |>
  arrange(desc(roc_auc)) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  head(5)

rf_tuned$results |>
  arrange(desc(roc_auc)) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

xgb_tuned$results |>
  arrange(desc(roc_auc)) |>
  select(nrounds, max_depth, roc_auc, sens, spec) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  head(5)
```




## ROC curves on validation set

```{r roc-curves}
library(pROC)

y_val_fac <- factor(val_prep$y, levels = c("No", "Yes"))

roc_logistic <- roc(y_val_fac, predict(fit_logistic, val_prep$x, type = "prob")[, "Yes"], quiet = TRUE)
roc_lasso    <- roc(y_val_fac, predict(fit_lasso, val_prep$x, type = "prob")[, "Yes"], quiet = TRUE)
roc_rf       <- roc(y_val_fac, predict(fit_rf, val_prep$x, type = "prob")[, "Yes"], quiet = TRUE)
roc_xgb      <- roc(y_val_fac, predict(fit_xgb, xgb.DMatrix(as.matrix(val_prep$x))), quiet = TRUE)

roc_df <- bind_rows(
  tibble(fpr = 1 - roc_logistic$specificities, tpr = roc_logistic$sensitivities, model = sprintf("Logistic (AUC=%.3f)", auc(roc_logistic))),
  tibble(fpr = 1 - roc_lasso$specificities,    tpr = roc_lasso$sensitivities,    model = sprintf("LASSO (AUC=%.3f)",    auc(roc_lasso))),
  tibble(fpr = 1 - roc_rf$specificities,        tpr = roc_rf$sensitivities,        model = sprintf("Random Forest (AUC=%.3f)", auc(roc_rf))),
  tibble(fpr = 1 - roc_xgb$specificities,       tpr = roc_xgb$sensitivities,       model = sprintf("XGBoost (AUC=%.3f)",  auc(roc_xgb)))
)

p_roc <- ggplot(roc_df, aes(x = fpr, y = tpr, color = model)) +
  geom_line(linewidth = 0.9) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "ROC Curves — Validation Set",
    x = "False Positive Rate (1 − Specificity)",
    y = "True Positive Rate (Sensitivity)",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p_roc)

ggsave("output/figures/13_roc_curves_validation.png", p_roc, width = 7, height = 6, dpi = 150)
```




## Feature importance (best model)

```{r feature-importance}
use_caret_importance <- best_model_name %in% c(
  "Logistic regression", "LASSO", "Random forest"
)

if (use_caret_importance) {
  best_model_obj <- switch(best_model_name,
    "Logistic regression" = fit_logistic,
    "LASSO"              = fit_lasso,
    "Random forest"      = fit_rf
  )
  imp <- varImp(best_model_obj, scale = TRUE)$importance |>
    rownames_to_column("feature") |>
    rename(importance = Overall) |>
    arrange(desc(importance)) |>
    slice_head(n = 20)

  p_imp <- ggplot(imp, aes(x = importance, y = reorder(feature, importance))) +
    geom_col(fill = "#4C78A8") +
    labs(
      title = paste("Top 20 Feature Importances —", best_model_name),
      x = "Scaled importance", y = NULL
    ) +
    theme_minimal()

  print(p_imp)
  ggsave("output/figures/14_feature_importance.png", p_imp, width = 8, height = 7, dpi = 150)
} else if (best_model_name == "XGBoost") {
  xgb_imp <- xgb.importance(model = fit_xgb) |>
    as_tibble() |>
    slice_head(n = 20)

  p_imp <- ggplot(xgb_imp, aes(x = Gain, y = reorder(Feature, Gain))) +
    geom_col(fill = "#4C78A8") +
    labs(
      title = "Top 20 Feature Importances — XGBoost",
      x = "Gain", y = NULL
    ) +
    theme_minimal()

  print(p_imp)
  ggsave("output/figures/14_feature_importance.png", p_imp, width = 8, height = 7, dpi = 150)
}
```




## Save trained models

```{r save}
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)

saveRDS(
  list(
    models = list(
      logistic = fit_logistic,
      lasso = fit_lasso,
      random_forest = fit_rf,
      xgboost = fit_xgb
    ),
    val_metrics = val_metrics_all,
    best_model = best_model_name,
    tuning = list(
      lasso = lasso_tuned$results,
      random_forest = rf_tuned$results,
      xgboost = xgb_tuned$results
    ),
    prep = prep_obj[c(
      "impute_from_train",
      "keep_predictors",
      "dummy_vars",
      "post_dummy_cols",
      "pre_process"
    )],
    train_ref = train_ref
  ),
  "data/processed/models.rds"
)
```